A Gaussian process model of human electrocorticographic data

### 2018-10-12 paper/main.tex
final edits through approach

The current ``gold standard'' for precisely localizing signals and
sampling at high temporal resolution is to take (ECoG or iEEG)
recordings from implanted electrodes (but from a limited set of
locations in any given brain).  This begs the following question: what
can we infer about the activity exhibited by the rest of a person's
brain, given what we learn from the limited intracranial recordings we
have from their brain and additional recordings taken from
\textit{other} people's brains?  Here we develop an approach, which we
call \textit{SuperEEG}\textit{SuperEEG}\footnote{The term ``SuperEEG''
  was coined by Robert J. Sawyer in his popular science fiction novel
  \textit{The Terminal Experiment}~\cite{Sawy95}.  SuperEEG is a
  fictional technology that measures ongoing neural activity
  throughout the entire living human brain with perfect precision and
  at arbitrarily high spatiotemporal resolution.}, based on Gaussian
process regression~\cite{Rasm06}.  SuperEEG entails using data from
multiple people to estimate activity patterns at arbitrary locations
in each person's brain (i.e., independent of their electrode
placements).  We test our SuperEEG approach using two large datasets
of intracranial recordings~\cite{SedeEtal03, SedeEtal07a, SedeEtal07b,
  MannEtal11, MannEtal12, EzzyEtal17, HoraEtal17, KragEtal17,
  KuceEtal17, LinEtal17, SoloEtal18, WeidEtal18, EzzyEtal18,
  KuceEtal18}.  We show that the SuperEEG algorithm recovers signals
well from electrodes that were held out of the training dataset.  We
also examine the factors that influence how accurately activity may be
estimated (recovered), which may have implications for electrode
design and placement in neurosurgical applications.

\section*{Approach}
The SuperEEG approach to inferring high temporal resolution full-brain
activity patterns is outlined and summarized in
Figure~\ref{fig:methods}. We describe (in this section) and evaluate
(in \textit{Results}) our approach using a two large previously
collected dataset comprising multi-session intracranial recordings.
Dataset 1 comprises multi-session recordings taken from 6876
electrodes implanted in the brains of 88 epilepsy
patients~\cite{SedeEtal03, SedeEtal07a, SedeEtal07b, MannEtal11,
  MannEtal12}.  Each recording session lasted from 0.2--3~h (total
recording time: 0.3--14.2~h; Fig.~\suppstats A).  During each
recording session, the patients participated in a free recall list
learning task, which lasted for up to approximately 1~h.  In addition,
the recordings included ``buffer'' time (the length varied by patient)
before and after each experimental session, during which the patients
went about their regular hospital activities (confined to their
hospital room, and primarily in bed).  These additional activities
included interactions with medical staff and family, watching
television, reading, and other similar activities.  For the purposes
of the Dataset 1 analyses presented here, we aggregated all data
across each recording session, including recordings taken during the
main experimental task as well as during non-experimental time.  We
used Dataset 1 to develop our main SuperEEG approach, and to examine
the extent to which SuperEEG might be able to generate task-general
predictions.  Dataset 2 comprised multi-session recordings from 4436
electrodes implanted in the brains of 40 epilepsy
patients~\citep{EzzyEtal17, HoraEtal17, KragEtal17, KuceEtal17,
  LinEtal17, SoloEtal18, WeidEtal18, EzzyEtal18, KuceEtal18}.  Each
recording session lasted from 0.4--2.2~h (total recording time:
0.4--6.6~h; Fig.~\suppstats B).  Whereas Dataset 1 included recordings
taken as the patients participated in a variety of activities, Dataset
2 included recordings taken as each patient performed each of two
specific experimental memory tasks: a random word list free recall
task (Experiment 1) and a categorized word list free recall task
(Experiment 2).  We used Dataset 2
to further examine the ability of SuperEEG to generalize its
predictions within versus across tasks.  Figure~\suppstats~provides
additional information about both datasets.

@@

We first applied fourth order Butterworth notch filter to remove 60 Hz
($\pm$ .5 Hz) line noise from every recording (from every electrode).
Next, we downsampled the recordings (regardless of the original
samplerate) to 250 Hz.  (This downsampling step served to both
normalize for differences in sampling rates across patients and to
ease the computational burden of our subsequent analyses.)  We then
excluded any electrodes that showed putative epileptiform activity.
Specifically, we excluded from further analysis any electrode that
exhibited an average kurtosis of 10 or greater across all of that
patient's recording sessions.  We also excluded any patients with
fewer than 2 electrodes that passed this criteria, as the SuperEEG
algorithm requires measuring correlations between 2 or more electrodes
from each patient.  For Dataset 1, this yielded clean recordings from
4168 electrodes implanted throughout the brains of 67 patients
(Fig.~\ref{fig:methods}A); for Dataset 2, this yielded clean
recordings from 3159 electrodes from 24 patients.  Each individual
patient contributes electrodes from a limited set of brain locations,
which we localized in a common space~\citep[MNI152;][]{GrabEtal06}; an
example Dataset 1 patient's 54 electrodes that passed the predefined
kurtosis test are highlighted in black and red.

@@

\begin{align}
\hat{K}_{\beta_s,\alpha_s} &= \hat{K}(\beta_s,\alpha_s),~\mathrm{and}\label{eqn:Kba}\\
\hat{K}_{\alpha_s,\alpha_s} &= \hat{K}(\alpha_s,\alpha_s)\label{eqn:Kaa}.
\end{align}
Here $\hat{K}_{\beta_s, \alpha_s}$ represents the correlations between
the ``unknown'' activity at the locations in $\beta_s$ and the
observed activity at the locations in $\alpha_s$, and
$\hat{K}_{\alpha_s, \alpha_s}$ represents the correlations between the
observed recordings (at the locations in $\alpha_s$).

@@

We designed our approach to be agnostic to electrode impedances, as
electrodes that do not exist do not have impedances.  Therefore our
algorithm recovers voltages in standard deviation ($z$-scored) units
rather than attempting to recover absolute voltages. (This property
reflects the fact that $\hat{K}_{\beta_s, \alpha_s}$ and
$\hat{K}_{\alpha_s, \alpha_s}$ are correlation matrices rather than
covariance matrices.)  Also, note that
Equation~\ref{eqn:reconstruction} requires computing a $T$ by $T$
matrix, which can become computationally intractable when $T$ is very
large (e.g., for the patient highlighted in Fig.~\ref{fig:recon}, $T =
12786750$). However, because Equation~\ref{eqn:reconstruction} is time
invariant, we may compute $Y_{s,k,\beta_s}$ in a piecewise
manner by filling in $Y_{s,k,\beta_s}$ one row at a time (using the
corresponding samples from $Y_{s, k, \alpha_s}$).

>>>>

The current ``gold standard'' for precisely localizing signals and
sampling at high temporal resolution is to take (ECoG or iEEG)
recordings from implanted electrodes (but from a limited set of
locations in any given brain).  This begs the following question: what
can we infer about the activity exhibited by the rest of a person's
brain, given what we learn from the limited intracranial recordings we
have from their brain and additional recordings taken from
\textit{other} people's brains?  Here we develop an approach, which we
call \textit{SuperEEG}\footnote{The term ``SuperEEG''
  was coined by Robert J. Sawyer in his popular science fiction novel
  \textit{The Terminal Experiment}~\cite{Sawy95}.  SuperEEG is a
  fictional technology that measures ongoing neural activity
  throughout the entire living human brain with perfect precision and
  at arbitrarily high spatiotemporal resolution.}, based on Gaussian
process regression~\cite{Rasm06}.  SuperEEG entails using data from
multiple people to estimate activity patterns at arbitrary locations
in each person's brain (i.e., independent of their electrode
placements).  We test our SuperEEG approach using two large datasets
of intracranial recordings~\cite{SedeEtal03, SedeEtal07a, SedeEtal07b,
  MannEtal11, MannEtal12, EzzyEtal17, HoraEtal17, KragEtal17,
  KuceEtal17, LinEtal17, SoloEtal18, WeidEtal18, EzzyEtal18,
  KuceEtal18}.  We show that the SuperEEG algorithm recovers signals
well from electrodes that were held out of the training dataset.  We
also examine the factors that influence how accurately activity may be
estimated (recovered), which may have implications for electrode
design and placement in neurosurgical applications.

\section*{Approach}
The SuperEEG approach to inferring high temporal resolution full-brain
activity patterns is outlined and summarized in
Figure~\ref{fig:methods}. We describe (in this section) and evaluate
(in \textit{Results}) our approach using a two large previously
collected dataset comprising multi-session intracranial recordings.
Dataset 1 comprises multi-session recordings taken from 6876
electrodes implanted in the brains of 88 epilepsy
patients~\cite{SedeEtal03, SedeEtal07a, SedeEtal07b, MannEtal11,
  MannEtal12}.  Each recording session lasted from 0.2--3~h (total
recording time: 0.3--14.2~h; Fig.~\suppstats E).  During each
recording session, the patients participated in a free recall list
learning task, which lasted for up to approximately 1~h.  In addition,
the recordings included ``buffer'' time (the length varied by patient)
before and after each experimental session, during which the patients
went about their regular hospital activities (confined to their
hospital room, and primarily in bed).  These additional activities
included interactions with medical staff and family, watching
television, reading, and other similar activities.  For the purposes
of the Dataset 1 analyses presented here, we aggregated all data
across each recording session, including recordings taken during the
main experimental task as well as during non-experimental time.  We
used Dataset 1 to develop our main SuperEEG approach, and to examine
the extent to which SuperEEG might be able to generate task-general
predictions.  Dataset 2 comprised multi-session recordings from 4436
electrodes implanted in the brains of 40 epilepsy
patients~\citep{EzzyEtal17, HoraEtal17, KragEtal17, KuceEtal17,
  LinEtal17, SoloEtal18, WeidEtal18, EzzyEtal18, KuceEtal18}.  Each
recording session lasted from 0.4--2.2~h (total recording time:
0.4--6.6~h; Fig.~\suppstats K).  Whereas Dataset 1 included recordings
taken as the patients participated in a variety of activities, Dataset
2 included recordings taken as each patient performed each of two
specific experimental memory tasks: a random word list free recall
task (Experiment 1) and a categorized word list free recall task
(Experiment 2).  We used Dataset 2
to further examine the ability of SuperEEG to generalize its
predictions within versus across tasks.  Figure~\suppstats~provides
additional information about both datasets.

@@

We first applied fourth order Butterworth notch filter to remove 60 Hz
($\pm$ .5 Hz) line noise from every recording (from every electrode).
Next, we downsampled the recordings (regardless of the original
samplerate) to 250 Hz.  (This downsampling step served to both
normalize for differences in sampling rates across patients and to
ease the computational burden of our subsequent analyses.)  We then
excluded any electrodes that showed putative epileptiform activity.
Specifically, we excluded from further analysis any electrode that
exhibited an average kurtosis of 10 or greater across all of that
patient's recording sessions.  We also excluded any patients with
fewer than 2 electrodes that passed this criteria, as the SuperEEG
algorithm requires measuring correlations between 2 or more electrodes
from each patient.  For Dataset 1, this yielded clean recordings from
4168 electrodes implanted throughout the brains of 67 patients
(Fig.~\ref{fig:methods}A); for Dataset 2, this yielded clean
recordings from 3159 electrodes from 24 patients.  Each individual
patient contributes electrodes from a limited set of brain locations,
which we localized in a common space~\citep[MNI152;][]{GrabEtal06}; an
example Dataset 1 patient's 54 electrodes that passed the above
kurtosis threshold test are highlighted in black and red.

@@

\begin{align}
\hat{K}_{\beta_s,\alpha_s} &= \hat{K}(\beta_s,\alpha_s),~\mathrm{and}\label{eqn:Kba}\\
\hat{K}_{\alpha_s,\alpha_s} &= \hat{K}(\alpha_s,\alpha_s)\label{eqn:Kaa}.
\end{align}
Here $\hat{K}_{\beta_s, \alpha_s}$ represents the correlations between
the ``unknown'' activity at the locations indexed by $\beta_s$ and the
observed activity at the locations indexed by $\alpha_s$, and
$\hat{K}_{\alpha_s, \alpha_s}$ represents the correlations between the
observed recordings (at the locations indexed by $\alpha_s$).

@@

We designed our approach to be agnostic to electrode impedances, as
electrodes that do not exist do not have impedances.  Therefore our
algorithm recovers voltages in standard deviation ($z$-scored) units
rather than attempting to recover absolute voltages. (This property
reflects the fact that $\hat{K}_{\beta_s, \alpha_s}$ and
$\hat{K}_{\alpha_s, \alpha_s}$ are correlation matrices rather than
covariance matrices.)  Also, we note that
Equation~\ref{eqn:reconstruction} requires computing a $T$ by $T$
matrix, which can become computationally intractable when $T$ is very
large (e.g., for the patient highlighted in Fig.~\ref{fig:recon}, $T =
12786750$). However, because Equation~\ref{eqn:reconstruction} is time
invariant, we may compute $Y_{s,k,\beta_s}$ in a piecewise
manner by filling in $Y_{s,k,\beta_s}$ one row at a time (using the
corresponding samples from $Y_{s, k, \alpha_s}$).
